### 2017-07-05 paper/prelims.tex
app

In this subsection we define the form of probabilistic programs that we 
consider in our analysis. We consider two classes of probabilistic programs: 
general probabilistic programs with arbitrary (measurable) expressions and 
their subclass, affine probabilistic programs (\APP s) where all expressions 
are restricted to be affine (see below for a precise definition). The reason 
for this dual view is that our work also has two main points of focus: a 
theoretical one, where we introduce new proof rules that can be used to prove 
properties of general probabilistic programs; and an algorithmic one, where we 
aim to prove properties of probabilistic programs automatically, using the 
aforementioned proof rules. As already testified in the non-probabilistic 
world, programs that contain only affine expressions allow for more efficient 
automation of the analysis and at the same time, due to the presence of 
non-determinism they can be used to form sound abstractions of programs with 
non-linear arithmetic. Hence, we consider general programs when providing our 
theoretical results and \APP s when presenting the automation of our techniques.

@@

\smallskip\noindent{\em Predicates.}
A \emph{predicate} is a logical formula obtained by a finite number of applications of conjunction, disjunction and negation operations on \emph{atomic predicates} of the form $E\leq E'$, where $E$, $E'$ are expressions.

@@

\textbf{[PETR: SAY SOMETHING ABOUT PREDICATE SATISFACTION.]}
\textbf{[PETR: EXPLAIN BOREL-MEASURABILITY IN MORE DETAIL?]}

@@

\smallskip\noindent{\em Affine Probabilistic Programs (\APP s).}
A probabilistic program is \emph{affine} if all the expressions that occur in 
the program (i.e. in loop guards, conditionals and on right-hand sides of 
assignments) are affine. We refer to the class of affine probabilistic programs 
as \APP s.
%We also assume that assume that each \APP{} $\program$ is preceded by an 
%initialization preamble 
%in 
%which 
%each variable appearing in $\program$ is assigned some concrete number.
%the program variables are initialized.

@@

\smallskip\noindent{\em Basics of Probability Theory.}
%%We now recall basic definitions of probability theory.
The crucial notion is the one of a probability space. A probability space is a triple
$(\Omega,\mathcal{F},\probm)$, where $\Omega$ is a non-empty set (so called
\emph{sample space}), $\mathcal{F}$ is a \emph{sigma-algebra} of measurable 
sets over $\Omega$,
i.e. a collection of subsets of $\Omega$ that contains the empty set
$\emptyset$, and that is closed under complementation and countable unions, and
$\probm$ is a \emph{probability measure} on $\mathcal{F}$, i.e., a function
$\probm\colon \mathcal{F}\rightarrow[0,1]$ such that
\begin{compactitem}
\item $\probm(\emptyset)=0$,
\item for all $A\in \mathcal{F}$ it holds $\probm(\Omega\smallsetminus
A)=1-\probm(A)$, and
\item for all pairwise disjoint countable set sequences $A_1,A_2,\dots \in
\mathcal{F}$ (i.e., $A_i \cap A_j = \emptyset$ for all $i\neq j$)
we have $\sum_{i=1}^{\infty}\probm(A_i)=\probm(\bigcup_{i=1}^{\infty} A_i)$.
\end{compactitem}

@@

\smallskip\noindent{\em Probabilistic Control Flow Graphs.}
%There are several ways in which one can express the semantics of \APP's with
%(angelic and demonic) 
%non-determinism
The semantics can be defined as the semantics of an uncountable state-space
Markov decision process (MDP) (uncountable due to real-valued variables).
We take an operational approach to define the semantics, and associate to each program 
a certain stochastic process~\cite{SriramCAV,HolgerPOPL,Kozen:prob-semantics}.
To define this process, we first define so called 
\emph{probabilistic control flow graphs}~\cite{CFG16:positivstellensatz-arxiv}.

\smallskip
\begin{definition}
\label{def:stochgame}
A \emph{probabilistic control flow graph (pCFG)} is a tuple
$\pCFG=(\locs,\pvars,\locinit,\vecinitset,\transitions,\updates,\probdist,\guards)$,
where
\begin{compactitem}
\item $\locs$ is a finite set of \emph{locations} partitioned into four 
pairwise
disjoint subsets  $\locsNB$, $\locsPB$, $\locs_D$, and $\locs_A$ of 
non-deterministic branching, 
probabilistic branching, deterministic, and assignment locations;
\item $\pvars=\{x_1,\dots,x_{|\pvars|}\}$ is a finite set of \emph{program 
variables} (note that $\pvars \subseteq \vars$) ;
\item $\locinit$ is an initial location and $\vecinitset$ is a set of initial 
\emph{assignment vectors};
% POKRACOVAT ODSUD
\item $\transitions\subseteq\locs\times\locs$ is a transition relation;
%whose members are tuples of
%the form $(\ell,i,\up,\ell')$, where $\ell$ and $\ell'$ are source and target
%program locations, respectively, $1\leq i \leq |\pvars|$ is a \emph{target 
%variable index}, and $u$ 
%is an 
%\emph{update element}, which can 
%be one of the following mathematical objects: %\begin{itemize}
%%\item 
%(a)~an affine function $u\colon \Rset^{|\pvars|}\rightarrow \Rset$;
%%\item 
%(b)~a distribution $d\in \mathcal{D}$; or
%%%\item 
%(c)~a set $R\subseteq \Rset$.
%\end{itemize}
\item $\updates$ is a function assigning to each transition outgoing from an 
assignment location a tuple $(i,\up)$, where $1\leq i \leq |\pvars|$ is a 
\emph{target 
	variable index} and $\up$ 
is an 
\emph{update element}, which can 
be one of the following mathematical objects: %\begin{itemize}
%\item 
(a)~a Borel-measurable function $u\colon \Rset^{|\pvars|}\rightarrow \Rset$;
%\item 
(b)~a distribution $d\in \mathcal{D}$; or
%\item 
(c)~a set $R\subseteq \Rset$.
%\end{itemize}
\item $\probdist=\{\prob_{\ell}\}_{\ell \in \locsPB}$ is a collection of
probability distributions, where each $\prob_{\ell}$ is a discrete probability
distribution on the set of all transitions outgoing from~$\ell$;
\item $\guards$ is a function assigning a propositionally linear predicate
(a \emph{guard}) over $\pvars$ to each transition outgoing from a deterministic 
location.
\end{compactitem}

We assume that each location has at least one outgoing transition.
Also, for every deterministic location $\ell$ we assume the following: if
$\tau_1,\dots,\tau_k$ are all transitions outgoing from $\ell$, then $G(\tau_1)
\vee \dots \vee G(\tau_k) \equiv \mathit{true}$ and $G(\tau_i) \wedge G(\tau_j)
\equiv \mathit{false}$ for each $1\leq i < j \leq k$. Moreover, for each 
distribution $d$ appearing in the 
pCFG we assume the following features are known: expected value $\expv[d]$ of 
$d$ and a 
single-variable PLP $\varphi_d$ such that the \emph{support} of $d$ (i.e. the 
smallest 
closed set of real numbers whose complement has probability zero 
under $d$)\footnote{In particular, a support of a \emph{discrete} probability 
	distribution $d$ is simply the at most countable set of all points on a 
	real 
	line that have positive probability under $d$.} satisfies 
$\support(d)\subseteq\exsat{\varphi_d}$. Finally, we assume that each 
assignment location has at most (and thus exactly) one outgoing transition.
%Finally, we assume that for each 
%transition $(\loc,j,u,\loc')$ such that $u$ is a set the location $\loc$ is 
%deterministic. This is just a technical assumption yielding no loss of 
%generality, and it somewhat simplifies notation.
%And we assume that each coordinate of $\mathcal{D}$ represents an integrable
%random variable (i.e., the expected value of the absolute value of the random
%variable exists).
\end{definition}

%We can naturally transfer the linear ordering of variables in $\vars$ to 
%variables in $X$. Hence, the set of variables $X$ can be uniquely written as a 
%vector $\vec{x}=(x_1,\dots,x_{|X|})$.
%Since $\pvars$ is a linearly ordered set of variables, there is a natural 
%correspondence between valuations in $\vals(\pvars)$ and 
%$|\pvars|$-dimensional 
%vectors. To formalize this correspondence, to any valuation 
%$\val\in\vals(X)$ we assign a vector 
%$\vec{x}_{\val}=(\val(x_1),\dots,\val(x_{|X|}))$, while to an 
%$|X|$-dimensional vector $\vec{x}$ we assign a valuation $\val_{\vec{x}}\in 
%\vals(\pvars)$ such that $\val_{\vec{x}}(x_i) = \vec{x}[i]$ for all $1\leq i 
%\leq |X|$.

%Given a pCFG $\pCFG$ 
%
%For notational convenience we assume that the sets $\pvars$ and $\rvars$ are
%endowed with some fixed linear ordering, which allows us to write $\pvars =
%\{x_1,x_2,\dots,x_{|\pvars|}\}$ and $\rvars = \{r_1,r_2,\dots,r_{|\rvars|}\}$.
%Every update function $f$ in a stochastic game can then be viewed as a tuple
%$(f_1,\dots,f_{|\pvars|})$, where each $f_i$ is of type
%$\Rset^{|\pvars\cup\rvars|}\rightarrow\Rset$. We denote by
%$\mathbf{x}=(\vec{x}[i])_{i=1}^{|\pvars|}$ and
%$\mathbf{r}=(\vec{r}[i])_{i=1}^{|\rvars|}$ the vectors of concrete valuations 
%of
%program and random variables, respectively.
%In particular, we assume that each component of $\vec{r}$ lies within the range
%of the corresponding random variable. %A valuation $\vec{r}$ is admissible if
%%each its component lies within the range of the corresponding random variable.
%We use the following succinct notation for special update functions: by $\id$ 
%we
%denote a function which does not change the program variables at all, i.e. for
%every $1\leq i \leq |\pvars|$ we have $f_i(\vec{x},\vec{r})= \vec{x}[i]$. For a
%function $g$ over the program and random variables we denote by 
%$\assgn{x_j}{g}$
%the update function $f$ such that $f_j(\vec{x},\vec{r})=g(\vec{x},\vec{r})$ and
%$f_i(\vec{x},\vec{r})=\vec{x}[i]$ for all $i\neq j$.

%We say that an SGS $\mathcal{G}$ is \emph{normalized} if all guards of all
%transitions in $\mathcal{G}$ are in a disjunctive normal form.

%\begin{example}\label{ex:sgs}
%Figure~\ref{fig:Q4:SGS} shows an example of stochastic game structure.
%Deterministic locations are represented by boxes, angelic locations by
%triangles, demonic locations by diamonds,
%and stochastic locations by circles. Transitions are labelled with update
%functions, while guards and probabilities
%of transitions outgoing from deterministic and stochastic locations,
%respectively, are given in rounded rectangles
%on these transitions. For the sake of succinctness we do not picture
%tautological guards and identity update functions. Note that the SGS is
%normalized.
%We will describe in Example~\ref{ex:illustrate} how the stochastic game
%structure shown corresponds to the program described in
%Example~\ref{ex:prog}.
%\end{example}

%\paragraph*{Dynamics of Stochastic Games}

The translation from probabilistic programs to the corresponding pCFG is 
standard~\cite{CFNH16:prob-termination}, and the details are presented 
in~\AppendixMaterial.

@@

\smallskip\noindent{\em Non-determinism and Schedulers.}
Due to the presence of non-determinism and probabilistic choices, a pCFG
$\pCFG$ may represent a multitude of possible behaviours. The probabilistic
behaviour of $\pCFG$ can be captured by constructing a suitable
probability measure over the set of all its runs. Before this can be
done, non-determinism in $\pCFG$ needs to be resolved. This is achieved using 
the 
standard notion of a \emph{scheduler}.

There are two sources of non-determinism in our programs: one in branching and 
one in assignments. We call a location $\loc$ non-deterministic if $\loc$ is a 
non-deterministic branching location or if $\loc$ is an assignment location 
with a 
non-deterministic assignment (i.e., the only transition $\tau$ outgoing from 
$\loc$ has $\updates(\tau)=(j,\up)$ where $\up$ is a set). A configuration 
$(\loc,\vec{x})$ is non-deterministic if $\loc$ is non-deterministic.

@@

The above intuitive explanation can be formalized by showing that
each pCFG $\pCFG$ together with a scheduler $\sigma$ and initial valuation 
$\vecinit\in \vecinitset$ uniquely determine a 
certain 
probabilistic 
space $(\OmegaRun,\natfilt,\probm^{\sigma}_{\vecinit})$ in which $\OmegaRun$ 
is a set of 
all 
runs in $\pCFG$, and a stochastic process 
$\pCFG^{\sigma}=\{\cfg{\sigma}{i}\}_{i=0}^{\infty}$ in this space 
such 
that for each run $\run\in \OmegaRun$ we have that $\cfg{\sigma}{i}(\run)$ is 
the 
$i$-th configuration on $\run$ (i.e., $\cfg{\sigma}{i}$ is a random vector 
$(\lr{\sigma}{i},\vr{\sigma}{i})$ with  $\lr{\sigma}{i}$ taking 
values in $\locs$ and $\vr{\sigma}{i}$ being a random vector of dimension 
$|\pvars|$ consisting of real-valued random variables). The sigma-algebra 
$\natfilt$ is the smallest (w.r.t. inclusion) sigma algebra under which all 
the functions $\cfg{\sigma}{i}$, for all $i\geq 0$, 
are $\natfilt$-measurable (i.e., for each $\cfg{\sigma}{i}$ and each measurable 
set of configurations $\confset\in\mathcal{F}_{\Conf}$ it holds 
$\{\run\mid\cfg{\sigma}{i}(\run)\in \confset\}\in \natfilt$). Equivalently, 
$\natfilt$ can be defined as a sigma algebra generated by all set of runs of 
the form $F\times \Conf^{\infty}$, where $F\in \mathcal{H}$ is a measurable set 
of finite paths. The 
probability 
measure 
$\probm^{\sigma}_{\vecinit}$ is such that for each $i$, the distribution of 
$\cfg{\sigma}{i}$ reflects the aforementioned way in which runs are randomly 
generated. The formal construction 
of $\natfilt$ and $\probm^{\sigma}_{\vecinit}$ proceeds via the standard 
\emph{cylinder 
construction}~\cite[Theorem 2.7.2]{Ash:book} and is
somewhat 
technical, hence we omit it. 
We denote by $\E^\sigma_{\vecinit}$ the expectation operator in probability 
space 
$(\OmegaRun,\natfilt,\probm^{\sigma}_{\vecinit})$.

@@

\vspace{-1em}
\subsection{Almost-Sure and Positive Termination}
\vspace{-0.5em}
We consider computational problems related to the basic liveness properties of 
\APP{s}, namely \emph{termination} and its generalization, \emph{reachability}.

\smallskip\noindent{\em Termination and termination time.}
In the following, consider a \PP{} $P$ and its associated pCFG $\pCFG_P$. 
We say that a run 
\emph{terminates} if it reaches a 
configuration whose first component (i.e. a location of $\pCFG_P$) is the 
location
$\loc_{P}^{\lout}$ corresponding to the value of the program counter after 
executing $P$. 
We define a random variable $\ttime$
such that for each run $\run$ the value $\ttime(\run)$ represents 
the first 
point in time when 
the current location is $\loc_P^{\lout}$. If a run $\run$ does 
\emph{not} 
terminate, then 
$\ttime(\run)=\infty$. We call $\ttime$ the 
\emph{termination time} of $\program$. 
%In particular, if $\confset$ is the set 
%of all configurations $(\loc,\vec{x})$ such that $\loc=\loc^\lout_P$ (the 
%terminal location of $\pCFG_P$), then $\treach{\confset}$ is 
%called a \emph{termination time}, as it returns the number of steps after 
%which 
%$\run$ terminates. Since termination time is an important concept on its own, 
%we use a special notation $\ttime$ for it.
%\begin{example}
%Consider the program $P$ such that \hspace{-0.25cm}\newsavebox{\figa}
%\lstset{language=affprob}
%\lstset{tabsize=1}
%\begin{lrbox}{\figa}
%\begin{lstlisting}[mathescape]
%$P=$x:=2;y:=1
%\end{lstlisting}
%\end{lrbox}
%\usebox{\figa}. The associated pCFG has three locations: initial location 
%$\loc_{P}^{\lin}$, terminal location $\loc_{P}^{\lout}$, and one 
%``intermediate'' location, say $\loc$. In the pCFG consider a run 
%$\run=(\loc_{P}^{\lin},(0,0))(\loc,(2,0))(\loc_{P}^{\lout},(2,1))(\loc_{P}^{\lout},(2,1))\cdots$.
% Then $\ttime(\run)=2$. If we denote by $C$ the set of configurations 
% $\{(\loc',(x,y))\mid \loc' \in\{\loc_{P}^{\lin},\loc_{P}^{\lout},\loc\} \text{ 
% and } x\geq 2\}$, then $\treach{C}(\run)=1$.
%\end{example}
Since a probabilistic program may exhibit more than one run, we are interested 
in probabilities of runs that terminate or reach some set of configurations. 
This gives rise to the following fundamental computational problems regarding 
termination:

>>>>

In this subsection we define the form of probabilistic programs that we 
consider in our analysis. We consider two classes of probabilistic programs: 
general probabilistic programs (\PP{}s ) with arbitrary (measurable) 
expressions and 
their subclass, affine probabilistic programs (\APP s) where all expressions 
are restricted to be affine (see below for a precise definition). The reason 
for this dual view is that our work also has two main points of focus: a 
theoretical one, where we introduce new proof rules that can be used to prove 
properties of general probabilistic programs; and an algorithmic one, where we 
aim to prove properties of probabilistic programs automatically, using the 
aforementioned proof rules. As already testified in the non-probabilistic 
world, programs that contain only affine expressions allow for more efficient 
automation of the analysis and at the same time, due to the presence of 
non-determinism they can be used to form sound abstractions of programs with 
non-linear arithmetic. Hence, we consider general programs when providing our 
theoretical results and \APP s when presenting the automation of our techniques.

@@

\smallskip\noindent{\em Predicates.}
A \emph{predicate} is a logical formula obtained by a finite number of 
applications of conjunction, disjunction and negation operations on 
\emph{atomic predicates} of the form $E\leq E'$, where $E$, $E'$ are 
expressions. We denote by $\vec{x}\models E$ the fact that $E$ is satisfied by 
substituting values from of $\vec{x}$ for the corresponding variables in $E$.

@@

%\textbf{[PETR: SAY SOMETHING ABOUT PREDICATE SATISFACTION.]}
%\textbf{[PETR: EXPLAIN BOREL-MEASURABILITY IN MORE DETAIL?]}

@@

\smallskip\noindent{\em Affine Probabilistic Programs (\APP s).}
A probabilistic program is \emph{affine} if all the expressions that occur in 
the program (i.e. in loop guards, conditionals, right-hand sides of 
assignments) are affine and if the set of possible initial valuations is a 
polyhedron. We refer to the class of affine probabilistic programs 
as \APP s.
%We also assume that assume that each \APP{} $\program$ is preceded by an 
%initialization preamble 
%in 
%which 
%each variable appearing in $\program$ is assigned some concrete number.
%the program variables are initialized.

@@

\smallskip\noindent{\em Basics of Probability Theory.}
%%We now recall basic definitions of probability theory.
 A probability space is a triple
$(\Omega,\mathcal{F},\probm)$, where $\Omega$ is a non-empty set (so called
\emph{sample space}), $\mathcal{F}$ is a \emph{sigma-algebra} of measurable 
sets over $\Omega$,
i.e. a collection of subsets of $\Omega$ that contains the empty set
$\emptyset$, and that is closed under complementation and countable unions, and
$\probm$ is a \emph{probability measure} on $\mathcal{F}$, i.e., a function
$\probm\colon \mathcal{F}\rightarrow[0,1]$ such that
\begin{compactitem}
\item $\probm(\emptyset)=0$,
\item for all $A\in \mathcal{F}$ it holds $\probm(\Omega\smallsetminus
A)=1-\probm(A)$, and
\item for all pairwise disjoint countable set sequences $A_1,A_2,\dots \in
\mathcal{F}$ (i.e., $A_i \cap A_j = \emptyset$ for all $i\neq j$)
we have $\sum_{i=1}^{\infty}\probm(A_i)=\probm(\bigcup_{i=1}^{\infty} A_i)$.
\end{compactitem}

@@

\smallskip\noindent{\em Probabilistic Control Flow Graphs.}
%There are several ways in which one can express the semantics of \APP's with
%(angelic and demonic) 
%non-determinism
We consider standard operational semantics of \PP{}s defined via an 
uncountable state-space
Markov decision process (MDP) (uncountable due to real-valued variables).
That is, we associate to each program 
a certain stochastic process.
To define this process, we first define so called 
\emph{probabilistic control flow graphs}~\cite{HONGFEI-CAV16}.

\smallskip
\begin{definition}
\label{def:stochgame}
A \emph{probabilistic control flow graph (pCFG)} is a tuple
$\pCFG=(\locs,\pvars,\locinit,\vecinitset,\transitions,\updates,\probdist,\guards)$,
where
\begin{compactitem}
\item $\locs$ is a finite set of \emph{locations} partitioned into four 
pairwise
disjoint subsets  $\locsNB$, $\locsPB$, $\locs_D$, and $\locs_A$ of 
non-deterministic branching, 
probabilistic branching, deterministic, and assignment locations;
\item $\pvars=\{x_1,\dots,x_{|\pvars|}\}$ is a finite set of \emph{program 
variables} (note that $\pvars \subseteq \vars$) ;
\item $\locinit$ is an initial location and $\vecinitset$ is a set of initial 
\emph{assignment vectors};
% POKRACOVAT ODSUD
\item $\transitions\subseteq\locs\times\locs$ is a transition relation;
%whose members are tuples of
%the form $(\ell,i,\up,\ell')$, where $\ell$ and $\ell'$ are source and target
%program locations, respectively, $1\leq i \leq |\pvars|$ is a \emph{target 
%variable index}, and $u$ 
%is an 
%\emph{update element}, which can 
%be one of the following mathematical objects: %\begin{itemize}
%%\item 
%(a)~an affine function $u\colon \Rset^{|\pvars|}\rightarrow \Rset$;
%%\item 
%(b)~a distribution $d\in \mathcal{D}$; or
%%%\item 
%(c)~a set $R\subseteq \Rset$.
%\end{itemize}
\item $\updates$ is a function assigning to each transition outgoing from an 
assignment location a tuple $(i,\up)$, where $1\leq i \leq |\pvars|$ is a 
\emph{target 
	variable index} and $\up$ 
is an 
\emph{update element}, which can 
be one of the following mathematical objects: %\begin{itemize}
%\item 
(a)~a Borel-measurable function $u\colon \Rset^{|\pvars|}\rightarrow \Rset$;
%\item 
(b)~a distribution $d\in \mathcal{D}$; or
%\item 
(c)~a set $R\subseteq \Rset$ (representing a non-deterministic update).
%\end{itemize}
\item $\probdist=\{\prob_{\ell}\}_{\ell \in \locsPB}$ is a collection of
probability distributions, where each $\prob_{\ell}$ is a discrete probability
distribution on the set of all transitions outgoing from~$\ell$;
\item $\guards$ is a function assigning a propositionally linear predicate
(a \emph{guard}) over $\pvars$ to each transition outgoing from a deterministic 
location.
\end{compactitem}

We assume that each location has at least one outgoing transition.
Also, for every deterministic location $\ell$ we assume the following: if
$\tau_1,\dots,\tau_k$ are all transitions outgoing from $\ell$, then $G(\tau_1)
\vee \dots \vee G(\tau_k) \equiv \mathit{true}$ and $G(\tau_i) \wedge G(\tau_j)
\equiv \mathit{false}$ for each $1\leq i < j \leq k$. For each 
distribution $d$ appearing in the 
pCFG we assume the following features are known: expected value $\expv[d]$ of 
$d$ and a set $SP_d$ containing the \emph{support} of $d$. Tthe support is the  
smallest 
closed set of real numbers whose complement has probability zero 
under $d$\footnote{In particular, a support of a \emph{discrete} probability 
	distribution $d$ is simply the at most countable set of all points on a 
	real 
	line that have positive probability under $d$.}) Finally, we assume that 
	each 
assignment location has at most (and thus exactly) one outgoing transition. The 
translation from probabilistic programs to the corresponding pCFG is 
standard~\cite{CFNH16:prob-termination}, and the details are presented 
in~\AppendixMaterial.
%Finally, we assume that for each 
%transition $(\loc,j,u,\loc')$ such that $u$ is a set the location $\loc$ is 
%deterministic. This is just a technical assumption yielding no loss of 
%generality, and it somewhat simplifies notation.
%And we assume that each coordinate of $\mathcal{D}$ represents an integrable
%random variable (i.e., the expected value of the absolute value of the random
%variable exists).
\end{definition}

@@

\smallskip\noindent{\em Non-determinism and Schedulers.}
%Due to the presence of non-determinism and probabilistic choices, a pCFG
%$\pCFG$ may represent a multitude of possible behaviours. 
The probabilistic
behaviour of $\pCFG$ can be captured by constructing a suitable
probability measure over the set of all its runs. Before this can be
done, non-determinism in $\pCFG$ needs to be resolved. This is achieved using 
the 
standard notion of a \emph{scheduler}. Note that there are two sources of 
non-determinism in our programs: one in branching and 
one in assignments. We call a location $\loc$ non-deterministic if $\loc$ is a 
non-deterministic branching location or if $\loc$ is an assignment location 
with the only transition $\tau$ outgoing from 
$\loc$ having a non-deterministic assignment. A configuration 
$(\loc,\vec{x})$ is non-deterministic if $\loc$ is non-deterministic.

@@

The above intuitive explanation can be formalized by showing that
each pCFG $\pCFG$ together with a scheduler $\sigma$ and initial valuation 
$\vecinit\in \vecinitset$ uniquely determine a 
certain 
probabilistic 
space $(\OmegaRun,\natfilt,\probm^{\sigma}_{\vecinit})$ in which $\OmegaRun$ 
is a set of 
all 
runs in $\pCFG$, and a stochastic process 
$\pCFG^{\sigma}=\{\cfg{\sigma}{i}\}_{i=0}^{\infty}$ in this space 
such 
that for each run $\run\in \OmegaRun$ we have that $\cfg{\sigma}{i}(\run)$ is 
the 
$i$-th configuration on $\run$ (i.e., $\cfg{\sigma}{i}$ is a random vector 
$(\lr{\sigma}{i},\vr{\sigma}{i})$ with  $\lr{\sigma}{i}$ taking 
values in $\locs$ and $\vr{\sigma}{i}$ being a random vector of dimension 
$|\pvars|$ consisting of real-valued random variables). The sigma-algebra 
$\natfilt$ is the smallest (w.r.t. inclusion) sigma-algebra under which all 
the functions $\cfg{\sigma}{i}$, for all $i\geq 0$, 
are $\natfilt$-measurable (i.e., for each $\cfg{\sigma}{i}$ and each measurable 
set of configurations $\confset\in\mathcal{F}_{\Conf}$ it holds 
$\{\run\mid\cfg{\sigma}{i}(\run)\in \confset\}\in \natfilt$). Equivalently, 
$\natfilt$ can be defined as a sigma algebra generated by all set of runs of 
the form $F\times \Conf^{\infty}$, where $F\in \mathcal{H}$ is a measurable set 
of finite paths. The 
probability 
measure 
$\probm^{\sigma}_{\vecinit}$ is such that for each $i$, the distribution of 
$\cfg{\sigma}{i}$ reflects the aforementioned way in which runs are randomly 
generated. The formal construction 
of $\natfilt$ and $\probm^{\sigma}_{\vecinit}$ proceeds via the standard 
\emph{cylinder 
construction}~\cite[Theorem 2.7.2]{Ash:book} and is
somewhat 
technical, hence we omit it. 
We denote by $\E^\sigma_{\vecinit}$ the expectation operator in probability 
space 
$(\OmegaRun,\natfilt,\probm^{\sigma}_{\vecinit})$.

@@

\vspace{-1em}
\subsection{Almost-Sure and Positive Termination}
\vspace{-0.5em}
Termination is the basic liveness property of \PP{}s.

\smallskip\noindent{\em Termination and termination time.}
In the following, consider a \PP{} $P$ and its associated pCFG $\pCFG_P$. This 
pCFG has a special location  $\loc_{P}^{\lout}$ corresponding to the value of 
the program counter after 
executing $P$.
We say that a run 
\emph{terminates} if it reaches a 
configuration whose first component is
$\loc_{P}^{\lout}$.
%
We define a random variable $\ttime$
such that for each run $\run$ the value $\ttime(\run)$ represents 
the first 
point in time when 
the current location is $\loc_P^{\lout}$. If a run $\run$ does 
\emph{not} 
terminate, then 
$\ttime(\run)=\infty$. We call $\ttime$ the 
\emph{termination time} of $\program$. 
%In particular, if $\confset$ is the set 
%of all configurations $(\loc,\vec{x})$ such that $\loc=\loc^\lout_P$ (the 
%terminal location of $\pCFG_P$), then $\treach{\confset}$ is 
%called a \emph{termination time}, as it returns the number of steps after 
%which 
%$\run$ terminates. Since termination time is an important concept on its own, 
%we use a special notation $\ttime$ for it.
%\begin{example}
%Consider the program $P$ such that \hspace{-0.25cm}\newsavebox{\figa}
%\lstset{language=affprob}
%\lstset{tabsize=1}
%\begin{lrbox}{\figa}
%\begin{lstlisting}[mathescape]
%$P=$x:=2;y:=1
%\end{lstlisting}
%\end{lrbox}
%\usebox{\figa}. The associated pCFG has three locations: initial location 
%$\loc_{P}^{\lin}$, terminal location $\loc_{P}^{\lout}$, and one 
%``intermediate'' location, say $\loc$. In the pCFG consider a run 
%$\run=(\loc_{P}^{\lin},(0,0))(\loc,(2,0))(\loc_{P}^{\lout},(2,1))(\loc_{P}^{\lout},(2,1))\cdots$.
% Then $\ttime(\run)=2$. If we denote by $C$ the set of configurations 
% $\{(\loc',(x,y))\mid \loc' \in\{\loc_{P}^{\lin},\loc_{P}^{\lout},\loc\} \text{ 
% and } x\geq 2\}$, then $\treach{C}(\run)=1$.
%\end{example}
Since a probabilistic program may exhibit more than one run, we are interested 
in probabilities of runs that terminate or reach some set of configurations. 
This gives rise to the following fundamental computational problems regarding 
termination:
